---
title: "Task 3 – Adverse Outcome Pathway (AOP) module of the STOP"
author: "Iris Martínez Hevia"
date: "May 15th 2024"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---
# Introduction
This Rmarkdown document provides a visualization of an Adverse Outcome Pathway (AOP) network retrieved from the AOP-Wiki using R package for igraph.

##  Installing required R packages
```{r message=FALSE, warning=FALSE}
if (!require("jsonlite")) install.packages("jsonlite")
if (!require("httr")) install.packages("httr")
if (!require("igraph")) install.packages("igraph")
```

##  Import libraries
```{r message=FALSE}
library(jsonlite)
library(httr)
library(igraph)
```

# Get AOP-wiki data
## Download of JSON for AOP#386
```{r}
#Get response
response <- httr::GET("https://aopwiki.org/aops/386.json", config = httr::config(ssl_verifypeer = FALSE))

# Extract content from the response
raw_data <- httr::content(response, as = "text")

# Parse JSON data into an R object
final_data <- fromJSON(raw_data)
```

## Extract information and visualization in a network
Information for events (MIE-molecular event, KE-key event and AO-Adverse outcome) and individual key event relationships (KERs) of AOP#386
```{r}
# Extracting relevant information
events <- rbind(final_data$aop_mies, final_data$aop_kes, final_data$aop_aos)
knitr::kable(events)

kers <- final_data$relationships
knitr::kable(kers)
```

# AOP#386 Network
This network is constructed using igraph package and a tree layout.
## Creating a graph object
```{r}
# Create a graph object
g <- graph(edges = c(), directed = TRUE)

# Set nodes for the different events
event_ids <- events$event_id
event_labels <- events$event
event_types <- events$event_type
event_colors <- ifelse(event_types == "KeyEvent", "dodgerblue", ifelse(event_types == "MolecularInitiatingEvent", "green", "red"))

# Add nodes
for (i in seq_along(event_ids)) {
  g <- add_vertices(g, nv = 1, name = event_ids[i], label = event_labels[i],
                    type = event_types[i], color = event_colors[i])
}

# Set edges for relationships (KERs)
ker_ids <- kers$relation
ker_sources <- kers$upstream_event_id 
ker_targets <- kers$downstream_event_id

#Add edges
for (i in seq_along(ker_ids)) {
  source_id <- as.character(ker_sources[i])
  target_id <- as.character(ker_targets[i])

    g <- add_edges(g, edges = c(source_id, target_id), color = "black")
}
```

## Creating a legend for the different nodes (Events)
```{r}
# Create a legend
# Extracting unique event types and their corresponding colors
unique_types <- unique(events$event_type)
type_colors <- ifelse(unique_types == "KeyEvent", "dodgerblue", ifelse(unique_types == "MolecularInitiatingEvent", "green", "red"))


legend_labels <- c("KE - Key Event", "MIE - Molecular Event", "AO - Adverse Outcome")
legend_colors <- type_colors
legend_text <- paste(legend_labels)
```


## Visualizing the network
```{r fig.height=12, fig.width=15}
par(cex.main = 2)

plot.igraph(g, 
            layout = layout_as_tree(g),
            vertex.label.cex = 2, 
            vertex.color = V(g)$color,
            vertex.size = 35,
            vertex.frame.color = "white",
            vertex.label.color = "black",
            vertex.label.font = 2,
            edge.arrow.size = 0.3,    
            edge.color = "black",
            main = final_data$short_name)

# Add legend
legend("bottomleft", legend = legend_text, fill = legend_colors, title = "Event Types", cex = 2)
```

# Integration of predicted PODs
## Load PoD data and PoD/AOP correspondences table
```{r}
# Load PoD
# The file is avaiable in the output folder of Task2_Assignment GitHub reposity
POD <- read.csv("/output/Result_Modeling.csv")

# Load correspondences between targets and event types:
correspondences <- data.frame("POD" = c("Reactive oxygen species", "Photosystem II efficiency",
                                        "Net photosynthetic rate", "Size", "Growth rate"),
                              "Event" = c("Increased, Reactive oxygen species", "Decrease, Photosystem II efficiency",
                                          "Decrease, Photosynthesis", "Decrease, Reproduction",
                                          "Decrease, Population growth rate"))
```

```{r}
# Merge the tables for creating the scheme for the network labels
colnames(POD)[3] <- "POD"
correspondences <- merge(correspondences, POD[,c("POD", "ED10")], by = "POD")
correspondences$ED10_round <- round(correspondences$ED10, 1)

knitr::kable(correspondences)
```

## Visualize final AOP network
```{r fig.height=12, fig.width=15}
par(cex.main = 2)

plot.igraph(g, 
            layout = layout_as_tree(g),
            vertex.label.cex = 2, 
            vertex.color = V(g)$color,
            vertex.size = 35,
            vertex.frame.color = "white",
            vertex.label.color = "black",
            vertex.label.font = 2,
            edge.arrow.size = 0.3,    
            edge.color = "black",
            main = final_data$short_name)

# Add legend
legend("bottomleft", legend = legend_text, fill = legend_colors, title = "Event Types", cex = 2)

# Add labels of different PoD to the AOP network
text(1, 0.4, paste("PoD for", correspondences[1,"POD"],"=", correspondences[1,"ED10_round"], "mGy/h"), cex = 1.5)
text(1, 0.90, paste("PoD for", correspondences[2,"POD"],"=", correspondences[2,"ED10_round"], "mGy/h"), cex = 1.5)
text(-1, 0.4, paste("PoD for", correspondences[3,"POD"],"=", correspondences[3,"ED10_round"], "mGy/h"), cex = 1.5)
text(1, -1.1, paste("PoD for", correspondences[4,"POD"],"=", correspondences[4,"ED10_round"], "mGy/h"), cex = 1.5)
```
